smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-08-29 manuscript/manuscript.qmd
implemented JC comments

Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR). While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms have to hold. We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.

@@

`smdi` enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe and infer potential missingness patterns and mechanisms based on available data.

@@

Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence coming from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments and lifestyle factors, EHRs can significantly improve the ability to control for confounding and imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often just partially observed which challenges the statistical analysis of the data and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

In light of these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`. To ensure the quality and robustness of the package, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment. Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

The workflow is generally categorized into descriptives, pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted. Examples are illustrated using a synthetic EHR dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent the variables relevant for the study, i.e., exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectivley. The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC. For a graphical visualization of this, the function returns a plot for each POC illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is (globally) MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would indicate the model is unable to discriminate missing vs. observed values based on available data; this could be due to a mechanism that is close to MCAR or one where the missingness is associated with unmeasured data (MNAR).

@@

The third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study. The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset. In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023]. As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC. Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.

@@

Finally, the `smdi_diagnose()` function enables a user to compute all three group diagnostics with just one function call.

@@

# minimal example of a smdi_diagnose() function call
smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3 # number of cores to parallelize computations on
  ) %>% 
  smdi_style_gt()
```

@@

The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data. Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.

@@

The smdi package is a powerful and convenient tool to implement and carry out principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.

>>>>

Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR). While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms must hold. We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.

@@

`smdi` enables users to run principled missing data investigations on partially observed confounders and implements functions to visualize, describe, and infer potential missingness patterns and mechanisms based on available data.

@@

Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their detailed capture of clinical parameters such as vital signs, lab measurements, physician assessments, and lifestyle factors, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when predicting or estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but general guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).

@@

To streamline implementation of routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`. To ensure quality, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment. Additional resources such as documentation and vignettes are provided on the package website: <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

The workflow is generally categorized into descriptives, pattern diagnostics, and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters users can specify, the returned results and how these results can be interpreted. Examples are illustrated using a synthetic dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns represent relevant variables, i.e., exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns, or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

To quantify such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMD along with a corresponding range of minimum and maximum AMSDs for each POC, respectivley. The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC. The function returns a plot for each POC illustrating the ASMD for each compared patient characteristic as a graphical visualization (example @fig-examples c) [@ggplot2].

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is (globally) MCAR.

@@

The group 2 diagnostics assess the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would indicate the model is unable to discriminate missing vs. observed values based on available data; this could be due to a mechanism that is close to MCAR or one where the missingness is associated with unmeasured data (MNAR).

@@

The third group of diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study. The function computes both a univariate model and a model adjusted for all other covariates in the dataset. In preceding simulations, we discerned distinct patterns in both univariate and adjusted associations between the missing indicator and the outcome, closely mirroring simulated missingness mechanisms [@weberpals2023]. As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC. Under MAR, given that missingness can be explained by observed covariates, a spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.

@@

Finally, the `smdi_diagnose()` function enables users to compute all three group diagnostics with a single function call.

@@

# minimal example of a smdi_diagnose() function call
smdi_diagnose(
  data = smdi_data,
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3
  ) %>% 
  smdi_style_gt()
```

@@

The package also has limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data. Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.

@@

The smdi package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.
